---
title: "Big 10 Graduate Student Stipend Analysis"
output: html_notebook
---

# Data preprocessing

```{r setup, message=FALSE}
library(tidyverse)
```

First, load in the self-reported stipend data. The school info is manually created, with COL data pulled from [here](https://livingwage.mit.edu/metros/14020).

```{r load_data, message=FALSE}
stipends <- 
  read_csv("http://www.phdstipends.com/csv") %>%
  janitor::clean_names() %>%
  
  # convert pays to numeric
  mutate(across(
    ends_with("_pay"),
    ~(
      str_remove_all(., pattern=",|\\$") %>%
        as.numeric()
    )  
  ))

# school-level information (COL, school names, etc.)
schools <- read_csv("school_data.csv")
```

Filter down data to the Big 10 and normalize the pay rates to the appropriate time periods. We also remove outliers: some responses are per-month or per-thousand, so we remove values below $9,000/year. Others include extraordinary fellowships and post-docs. These assumptions would have to be re-evaluated if looking at the international schools.

```{r cleanup}
big10 <-
  stipends %>%
  filter(university %in% schools$university) %>%
  mutate(
     # 3-month data too noisy to consider
     pay_12_or_9 = coalesce(x12_m_gross_pay, x9_m_gross_pay)
  ) %>%
  
  # remove outliers
  filter(pay_12_or_9 > 9000 & pay_12_or_9 < 75000) %>%
  
  # create a 9-month & 12-month equivalent rate
  mutate(
    # counts only academic year
    pay_9_equivalent = coalesce(x9_m_gross_pay, x12_m_gross_pay * (3 / 4)),
    
    # does not include summer funding, since this is a bit noisy and may not be
    # guaranteed
    pay_12_no_summer = coalesce(x12_m_gross_pay, x9_m_gross_pay),
    
    # includes summer funding
    pay_12_plus_summer = coalesce(
      x12_m_gross_pay,
      x9_m_gross_pay + coalesce(x3_m_gross_pay, 0)
    ),
  )
```

# Create school summaries

First, create a function that can calculate the Cost-of-Living (COL) deficit, either as a ratio or a difference.
```{r col_deficit_fun}
calc_col_deficit <- function(df, fn=`/`){
  df %>%
    mutate(
      col_9_mo = col * (3 / 4),
      pay_9_to_col = fn(pay_9_equivalent, col_9_mo),
      pay_12_no_summer_to_col = fn(pay_12_no_summer, col),
      pay_12_plus_summer_to_col = fn(pay_12_plus_summer, col),
      min_stipend_to_col = fn(min_stipend, col)
    )
}
```

Then, group our stipend data by school and calculate summary statistics (e.g., means or medians) of the pay data for each school.
```{r summarize, message=FALSE}
# Set the averaging function (e.g., mean or median) and the 
# COL deficit function (e.g,, - or / )
avg_fn <- median
col_def_fn <- `-`

big10_counts <- group_by(big10, university) %>% summarize(responses=n())

big10_summary <-
  big10 %>%
  group_by(university) %>%
  summarize(across(where(is.numeric), avg_fn, na.rm=TRUE)) %>%
  left_join(big10_counts) %>%
  left_join(schools) %>%
  calc_col_deficit(col_def_fn) %>%
  select(
    university, uni, is_unionized, responses,
    starts_with("col"), starts_with("pay"), min_stipend, min_stipend_to_col, -pay_12_or_9
  )

write_csv(big10_summary, "big10_summary.csv")
```

For posterity, here is column info:
```
Columns:
    university:              University name
    is_unionized:            Is campus unionized (includes Northwestern, which is ongoing)
    responses:               Individual survey responses in data
    col:                     Required annual cost of living, per MIT data
    col_9_mo:                col*(3/4)
    pay_9_equivalent:        9-month pay if reported, otherwise reported 12-month pay multiplied by 3/4
    pay_12_no_summer:        12-month pay if reported, otherwise flat 9-month pay
    pay_12_plus_summer:      12-month pay if reported, otherwise 9-month pay plus any reported 3-month pay (may include outside internships; data are noisy)
    min_stipend:             minimum stipend reported for the school
    <payment_column>_to_col: ratio of pay to COL over the specified period
```

# Analyze the data

```{r }
```

# Plot the

```{r big10_deficit_plot_fn}
plot_big10_deficits <- function(df, pay_variable="pay_9_to_col"){
  pay_period <- c(
    pay_9_to_col="(median, academic year)",
    pay_12_no_summer_to_col="(median, full-year)",
    pay_12_plus_summer_to_col="(median, full-year w/ summer funding)",
    min_stipend_to_col=" (based on minimum stipend)"
  )
  
  df %>%
    mutate(
      union_labels = if_else(is_unionized, "Unionzied", "Not Unionized"),
      union_labels = factor(
        if_else(uni == "UMD", "UMD", union_labels),
        levels=c("Unionzied", "Not Unionized", "UMD"),
        ordered=TRUE
      )
    ) %>%
    arrange(is_unionized, .data[[pay_variable]]) %>%
    mutate(
      order = 1:n(),
    ) %>%
    ggplot(aes(
          x = reorder(uni, order),
          y = !!ensym(pay_variable),
    )) +
    geom_bar(aes(fill=union_labels), stat="identity") +
    scale_fill_manual(
      values=c("#ffd200", "gray", "#E21833"),
    ) +
    scale_y_continuous(labels=scales::dollar_format()) +
    labs(
      title = paste("Cost of Living Deficit by University", pay_period[pay_variable], sep="\n"),
      x = "",
      y = paste("Cost-of-Living Deficit", pay_period[pay_variable], sep="\n"),
      fill = ""
    ) +
    theme_classic() + 
    theme(
      text = element_text(size=12),
      axis.text.x = element_text(angle=45, hjust=1)
    )
}
```

```{r big10_deficit_plots}
plot_big10_deficits(big10_summary, pay_variable = "pay_9_to_col")
plot_big10_deficits(big10_summary, pay_variable = "pay_12_plus_summer_to_col")
plot_big10_deficits(big10_summary, pay_variable = "min_stipend_to_col")
```